WHOOMP! (There It Is) \\ Rapid Bayesian position reconstruction for gravitational\nobreakdashes-wave transients

### 2014-05-22 ms.tex
Some minor edits

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques is several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
	x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[n_i(\omega)^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

@@

Bayesian parameter estimation has many advantages, including broad generality and the ability to make probabilistically meaningful statements even with very low \ac{SNR} measurements. However, in problems of even modest complexity, the marginalization step burdens the analysis with nasty, many-dimensional, ill-behaved integrals. Almost all modern parameter estimation tactics are variations of \ac{MCMC} integration. Though powerful, \ac{MCMC} is slow and non-deterministic. With the most sophisticated \ac{CBC} parameter estimation codes, it still takes days to process a single event. This delay is undesirable for planning targeted \ac{EM} follow\nobreakdashes-up searches of \ac{LIGO} events.

@@

The first task is to understand the uncertainty in the \ac{ML} point estimates from the detection. We begin with the \ac{CRLB}, which constraints the covariance of any unbiased estimator. The \ac{CRLB} has been widely applied in \ac{GW}. Here, we review the Fisher matrix calculation in \cite{fairhurst:2009} and discuss the interpretation.

@@

First, we put the phase and time on the same footing by measuring the time in units of $1 / 2 \pi \sqrt{\overline{\omega^2}}$ with a change of variables from $\tau$ to $\gamma_\tau = 2 \pi \sqrt{\overline{\omega^2}} \tau$:
%
\begin{equation}
    \mathcal{I}' = \bordermatrix{
        ~ & \rho_i & \gamma_i & \gamma_{\tau,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} \cr
        \gamma_{\tau,i} & 0 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} & {\rho_i}^2
    }.
\end{equation}
%
The second change of variables, from $\gamma$ and $\gamma_\tau$ to $\gamma_\pm = \frac{1}{\sqrt{2}}(\gamma \pm \gamma_\tau)$, diagonalizes the Fisher matrix:
%
\begin{equation}
    \mathcal{I}'' = \bordermatrix{
        ~ & \rho_i & \gamma_{+,i} & \gamma_{-,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_{+,i} & 0 & \left(1 - \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2 & 0 \cr
        \gamma_{-,i} & 0 & 0 & \left(1 + \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2
    }.
\end{equation}
%
Thus, in the appopriate time units, the \textit{sum and difference} of the phase and time of the signal are measured independently.

@@

There are many ways to realize a distribution function with these properties, but they tend to be difficult to work with.%
%
\marginpar{Give some examples.}%
%
An alternative is to just discard the one periodic parameter, $\gamma$, entirely. We do not know the exact distribution of the remaining parameters $\rho$ and $\tau$, but we are less liable to make egregious errors by assuming that they are Gaussian than if we assumed that the estimation errors in $\gamma$ were Gaussian. Projecting out $\gamma$ gives us the covariance matrix
%
\begin{equation}
    \cov \left(
        \begin{array}{c}
            \hat{\rho} - \rho \\
            \hat{\tau} - \tau
        \end{array}
    \right) =
    \left(
        \begin{array}{cc}
            1 & 0 \\
            0 & \rho^2/{\omega_\mathrm{rms}}^2
        \end{array}
    \right).
\end{equation}

@@

\section*{References}
\bibliographystyle{iopart-num}
\bibliography{apj-jour,bib/telescope}

>>>>

In this paper we present a rapid and accurate method of sky localization that makes use of Bayesian methods. It differs from existing techniques in several important ways.  In the first place, we fix the intrinsic parameters to their \ac{ML} estimates, as provided by the detection pipeline.  This reduces the dimensionality of the parameter space we need to sample. In addition, the technique takes as its input the \ac{ML} estimates of the extrinsic parameters, instead of the $h(t)$ time series. This makes the likelihood itself much faster to evaluate. Finally, instead of of using \ac{MCMC} or some similar method for statistical sampling, we make use of a deterministic quadrature scheme. We call this technique \ac{BAYESTAR}%
%
\footnote{A pun on the Cylon battleships in the American television series Battlestar Galactica. The defining characteristic of the Cylons is that they repeatedly defeat humanity by using their superhuman information\nobreakdashes-gathering ability to coordinate overwhelming forces. The name also suggests that, like the Cylons, \ac{GW} detectors may some day rise against us humans.}%
%
\footnote{We do not like to mention the final `L' in the acronym, because then it would be pronounced BAYESTARL, which sounds stupid.}%
. It is unique in that in bridges the detection and parameter estimation of \ac{GW} signals, two tasks that have until now involved very different numerical methods and time scales. We expect that \ac{BAYESTAR} will take on a key role in observing \ac{CBC} events in both \ac{GW} and optical channels during the Advanced \ac{LIGO} era.

@@

In the frequency domain, the strain observed by a single \ac{GW} interferometer is
%
\begin{equation}\label{eq:signal-model}
	x_i (\omega) = h_i (\omega; \boldsymbol\theta) + n_i (\omega),
\end{equation}
%
where $h_i (\omega; \boldsymbol\theta)$ is the \ac{GW} signal given a parameter vector $\boldsymbol\theta$ that describes the \ac{GW} source, and $n_i (\omega)$ is that detector's \ac{WSS} Gaussian noise with one\nobreakdashes-sided \ac{PSD} $S_i(\omega) = E\left[\left|n_i(\omega)\right|^2\right]$. We shall denote the combined observation from a network of detectors as $\mathbf x (\omega) \equiv \{x_i (\omega)\}_i$.

@@

Bayesian parameter estimation has many advantages, including broad generality and the ability to make probabilistically meaningful statements even with very low \ac{SNR} measurements. However, in problems of even modest complexity, the marginalization step burdens the analysis with nasty, many-dimensional, ill-behaved integrals. Almost all modern parameter estimation tactics are variations of \ac{MCMC} integration. Though powerful, \ac{MCMC} is historically slow and non-deterministic. With the most sophisticated \ac{CBC} parameter estimation codes, it still takes days to process a single event. This delay is undesirable for planning targeted \ac{EM} follow\nobreakdashes-up searches of \ac{LIGO} events.

@@

The first task is to quantify the uncertainty in the \ac{ML} estimates of the \acp{TOA} and amplitudes from the detection. We begin with the \ac{CRLB}, which constraints the covariance of any unbiased estimator. The \ac{CRLB} has been widely applied in \ac{GW}. Here, we review the Fisher matrix calculation in \cite{fairhurst:2009} and discuss the interpretation.

@@

First, we put the phase and time on the same footing by measuring the time in units of $1 / \sqrt{\overline{\omega^2}}$ with a change of variables from $\tau$ to $\gamma_\tau = \sqrt{\overline{\omega^2}} \tau$:
%
\begin{equation}
    \mathcal{I}' = \bordermatrix{
        ~ & \rho_i & \gamma_i & \gamma_{\tau,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_i & 0 & {\rho_i}^2 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} \cr
        \gamma_{\tau,i} & 0 & -{\rho_i}^2\frac{{\overline{\omega}}_i}{\sqrt{{\overline{\omega^2}}_i}} & {\rho_i}^2
    }.
\end{equation}
%
The second change of variables, from $\gamma$ and $\gamma_\tau$ to $\gamma_\pm = \frac{1}{\sqrt{2}}(\gamma \pm \gamma_\tau)$, diagonalizes the Fisher matrix:
%
\begin{equation}
    \mathcal{I}'' = \bordermatrix{
        ~ & \rho_i & \gamma_{+,i} & \gamma_{-,i} \cr
        \rho_i & 1 & 0 & 0 \cr
        \gamma_{+,i} & 0 & \left(1 - \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2 & 0 \cr
        \gamma_{-,i} & 0 & 0 & \left(1 + \frac{\overline{\omega}_i}{\sqrt{\overline{\omega^2}_i}}\right){\rho_i}^2
    }.
\end{equation}
%
Thus, in the appopriate time units, the \textit{sum and difference} of the phase and time of the signal are measured independently.

\subsection{Barankin bound}

It is well known that the \ac{CRLB} is not a tight bound in the sense that there does not always exist an estimator that achieves it. Barankin derived a tight bound for any unbiased estimator \cite{barankin1949locally}.

@@

There are many ways to realize a distribution function with these properties, but they tend to be difficult to work with.
%
\marginpar{Give some examples.}%
%
An alternative is to just discard the one periodic parameter, $\gamma$, entirely. We do not know the exact distribution of the remaining parameters $\rho$ and $\tau$, but we are less liable to make egregious errors by assuming that they are Gaussian than if we assumed that the estimation errors in $\gamma$ were Gaussian. Projecting out $\gamma$ gives us the covariance matrix
%
\begin{equation}
    \cov \left(
        \begin{array}{c}
            \hat{\rho} - \rho \\
            \hat{\tau} - \tau
        \end{array}
    \right) =
    \left(
        \begin{array}{cc}
            1 & 0 \\
            0 & \rho^2/{\omega_\mathrm{rms}}^2
        \end{array}
    \right).
\end{equation}

@@

\section*{References}
\bibliographystyle{iopart-num}
\bibliography{apj-jour,bib/telescope,ms}
